Formalizing Stack Safety as a Security Property

### 2022-11-15 submission5/noexceptions.tex
Most updates

Collectively addresses and registers are {\em state elements} \(\component\)
in the set \(\COMPONENTS ::= \WORDS + \REGS + \mathcal{P}\).
%
The set of {\em machine states} \(\MACHS\) (ranged over by \(\mach\) and \(\nach\))
defines the mapping \(\mach[\component]\) from state elements to words defined.

@@

While some enforcement mechanisms may cause a machine to
failstop, we model this as the machine silently diverging by remaining in the
same state. Since failstops are not distinguished from normal execution,
our properties will suffer the limitations typical of
{\it termination insensitive} systems \cite{}. TODO: more detailed explanation later.

@@

Security-relevant operations consist of calls, returns, and allocations.
Calls identify the address of the call's target,
and the argument registers. Allocations and deallocations are phrased in terms of a pair
of an offset from the stack pointer and a size. Returns don't require any additional information.
%
\begin{align*}
  \psi \in \Psi ::= & \mathbf{call} ~ \addr_{target} ~ \overline{\reg_{args}} & \\
  | & \mathbf{return} & \\
  | & \mathbf{alloc} ~ \mathit{off ~ sz} & \mathit{off, sz} \in \mathbb{N} \\
\end{align*}
%
Security principals are function activations: at any given time, pending activations
have security requirements that are recorded in a call stack, and the current
activation tracks context information that will inform its requirements when
it makes a call. Each activation has a {\it view}
of the system that maps each state element to a {\it security class}:
it may be private to an inactive principal (\(\sealed\)),
in an allocated object that is accessible to the current principal (\(\object\)),
available to be allocated (\(\unsealed\)), or
always accessible (\(\public\)). The {\it initial view} \(V_0\) maps all stack locations to \(\unsealed\),
all other locations to \(\public\), and registers based on which set they
belong to (Figure \ref{fig:RISCVregs}): \(\sealed\) for callee-saved,
\(\unsealed\) for caller-saved, and \(\public\) otherwise. Elements of \(\mathcal{P}\),
such as the program counter, are always \(\public\).

@@

\begin{figure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{lw a4,8(sp)}{}
      \labeledrow{104:}{sw a4,out}{}
      \labeledrow{108:}{li a0,1}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \caption{Leaking {\tt secret} directly}
    \label{subfig:direct}
  \end{subfigure}  
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{lw a4,8(sp)}{}
      \labeledrow{104:}{mov a4,a0}{}
      \labeledrow{108:}{sw zero,-4(sp)}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \caption{Leaking {\tt secret} indirectly}
    \label{subfig:indirect}
  \end{subfigure}  
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{li a5,42}{}
      \labeledrow{104:}{sw a5,4(sp)}{}
      \labeledrow{108:}{li a0,1}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \subcaption{Attacking {\tt sensitive}}
    \label{subfig:integrity}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{addi ra,ra,16}{}
      \labeledrow{104:}{nop}{}
      \labeledrow{108:}{nop}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \subcaption{Attacking control flow}
    \label{subfig:WBCF}
  \end{subfigure}

@@

    \includegraphics[width=\columnwidth]{stack1.png}
    \includegraphics[width=\columnwidth]{stack2.png}
    \includegraphics[width=\columnwidth]{stack3.png}
    \includegraphics[width=\columnwidth]{stack4.png}
  \end{subfigure}

@@

To see why WBCF also cares about the stack pointer, consider an attack \apt{should this
  be shown as a fifth attack in the figure?} which
returns with \(\SP' = 998\) instead of the correct \(\SP = 980\), and recall
the layout of {\tt main}'s frame: 
\begin{tabular}{| l | l | l | l | l |}
  \multicolumn{1}{r}{\(\SP \downarrow\)} &
  \multicolumn{2}{r}{\(\SP' \downarrow\)} \\
  \hline
  res & sens & sec & ra & ra \\
  \hline
\end{tabular}

@@

Like WBCF, stack integrity defines at the call a condition that must hold upon
return. This time the condition applies to all of the memory that a function has
allocated. In Figure \ref{fig:exec1} we see the lifecycle of an allocated frame:
upon allocation, the view labeled is \(\object\), and when a call is made, it instead
becomes \(\sealed\). Intuitively, the integrity of {\tt main}
is preserved if, when control returns to it, it can rely on any \(\sealed\) elements
to be identical to when it made the call.
%
Again, we need to know when a caller has been returned to,
and we use the same mechanism of checking the depth of the call stack.
%
In the case of the call from {\tt main} to {\tt f}, the \(\sealed\) elements are the
addresses 980 through 999 and callee-saved registers such as
the stack pointer. Note that callee-saved registers often change
during the call---but if so, the callee is obligated to restore them to their prior value.

While this notion of integrity is simple, it is stricter than necessary for many purposes.
For example, it arguably doesn't matter if {\tt f} overwrites {\tt main}'s data provided that {\tt main} never
reads that corrupted data.  This is the justification behind the lazy micro-policy~\cite{??}:
the monitor permits the caller to write into the
callee's frame, but taints that write so that the callee cannot read it later without
triggering a failstop. More generally, corruption by {\tt f} can be considered acceptable
as long as it does not affect {\tt main}'s observable behavior. \apt{Maybe justify this
  point of view by appeal to standard security notions?}
To characterize this situation, we define a state element to be \emph{irrelevant}
if its value can be replaced with any other value without changing the observable behavior
of the machine. More precisely, for a set of elements \(\components\),
a pair of states \(\mach\) and \(\nach\)
are {\em \(\components\)-variants} if they agree on the value of every
element except those in \(\components\), and share the same policy state\apt{hopefully can remove this last}.
The elements of \(\components\) are \emph{irrelevant} in \(\mach\) if every variant state produces an equivalent trace
(see Section \ref{sec:events}); all other elements are \emph{relevant}.
We characterize \emph{observational integrity} as the property that
every relevant element that is \(\sealed\) under the callee's view is restored
to its original value at the return point.

>>>>

Collectively addresses and registers are {\em state elements} \(\component\)
in the set \(\COMPONENTS ::= \WORDS + \REGS + \mathcal{P}\). A {\em machine state}
\(\mach\) is a map from components to words, written \(\mach[\component]\).

\sna{It occurs to me, based on some of your comments further down, that the old machine
  states were unclear. My intention was not for \(\mach\) to be a map, but for
  it to be an arbitrary type on which \(\cdot[\cdot]\) was defined. (Akin to
  the coq idiom of a record with a field t:Type and proj:t\(\rightarrow\)k\(\rightarrow\)w.)
  Therefore, implicitly, there could be additional state for enforcement and other
  purposes. The new presentation with P makes that explicit but is less elegant.
}

@@

While some enforcement mechanisms may cause a machine to
failstop, we model this as the machine silently diverging by remaining in the
same state. Since failstops are not distinguished from normal divergence,
our properties will suffer the limitations typical of
{\it termination insensitive} systems \cite{}. TODO: more detailed explanation later.

@@

Security-relevant operations consist of calls, returns, and allocations.
Calls identify the address of the call's target,
and the argument registers. Allocations and deallocations are phrased in terms of a pair
of an offset from the stack pointer and a size. Returns don't require any additional information.
%
\begin{align*}
  \psi \in \Psi ::= & \mathbf{call} ~ \addr_{target} ~ \overline{\reg_{args}} & \\
  | & \mathbf{return} & \\
  | & \mathbf{alloc} ~ \mathit{off ~ sz} & \mathit{off, sz} \in \mathbb{N} \\
\end{align*}
%
Security principals are function activations: at any given time, pending activations
have security requirements that are recorded in a call stack, and the current
activation tracks context information that will inform its requirements when
it makes a call. Each activation has a {\it view}
of the system that maps each state element to a {\it security class}:
it may be private to an inactive principal (\(\sealed\)),
in an allocated object that is accessible to the current principal (\(\object\)),
available to be allocated (\(\unsealed\)), or
always accessible (\(\public\)). The {\it initial view} \(V_0\) maps all stack locations to \(\unsealed\),
all other locations to \(\public\), and registers based on which set they
belong to (Figure \ref{fig:RISCVregs}): \(\sealed\) for callee-saved,
\(\unsealed\) for caller-saved, and \(\public\) otherwise. Elements of \(\mathcal{P}\),
such as the program counter, are always \(\public\). Even if they might deserve their
own protection, that protection is orthogonal to stack safety.

\sna{And the payoff of \(\mathcal{P}\) being explicit: we can just state this.}

@@

\begin{figure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{lw a4,8(sp)}{}
      \labeledrow{104:}{sw a4,out}{}
      \labeledrow{108:}{li a0,1}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \caption{Leaking {\tt secret} directly}
    \label{subfig:direct}
  \end{subfigure}  
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{lw a4,8(sp)}{}
      \labeledrow{104:}{mov a4,a0}{}
      \labeledrow{108:}{sw zero,-4(sp)}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \caption{Leaking {\tt secret} indirectly}
    \label{subfig:indirect}
  \end{subfigure}  
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{li a5,42}{}
      \labeledrow{104:}{sw a5,4(sp)}{}
      \labeledrow{108:}{li a0,1}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \subcaption{Attacking {\tt sensitive}}
    \label{subfig:integrity}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{addi ra,ra,16}{}
      \labeledrow{104:}{nop}{}
      \labeledrow{108:}{nop}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \subcaption{Attacking control flow}
    \label{subfig:WBCF}
  \end{subfigure}
  \begin{subfigure}[b]{\columnwidth}
    \vspace{\abovedisplayskip}
    \begin{tabular}{r l | l}
      \labeledrow{100:}{addi sp,sp,8}{}
      \labeledrow{104:}{nop}{}
      \labeledrow{108:}{nop}{}
      \labeledrow{112:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
    \subcaption{Relative address attack}
    \label{subfig:WBCF2}
  \end{subfigure}

@@

    \(\underbrace{~ ~ ~}_\unsealed \uparrow \SP \)

    \includegraphics[width=\columnwidth]{stack2.png}
    \includegraphics[width=\columnwidth]{stack3.png}
    \includegraphics[width=\columnwidth]{stack4.png}
  \end{subfigure}

@@

To see why WBCF also cares about the stack pointer, consider the attack in
Figure \ref{subfig:WBCF2}. Here the attacker returns with \(\SP' = 998\) instead of the
correct \(\SP = 980\). Given the layout of {\tt main}'s frame: 
\begin{tabular}{| l | l | l | l | l |}
  \multicolumn{1}{r}{\(\SP \downarrow\)} &
  \multicolumn{2}{r}{\(\SP' \downarrow\)} \\
  \hline
  res & sens & sec & ra & ra \\
  \hline
\end{tabular}

@@

Like WBCF, stack integrity defines at the call a condition that must hold upon
return. This time the condition applies to all of the memory that a function has
allocated. In Figure \ref{fig:exec1} we see the lifecycle of an allocated frame:
upon allocation, the view labeled is \(\object\), and when a call is made, it instead
becomes \(\sealed\). Intuitively, the integrity of {\tt main}
is preserved if, when control returns to it, it can rely on any \(\sealed\) elements
to be identical to when it made the call.
%
Again, we need to know when a caller has been returned to,
and we use the same mechanism of checking the depth of the call stack.
%
In the case of the call from {\tt main} to {\tt f}, the \(\sealed\) elements are the
addresses 980 through 999 and callee-saved registers such as
the stack pointer. Note that callee-saved registers often change
during the call---but if the caller accesses them after the call, it should find them
restored to their prior value.

While it would be simple to define integrity as ``all sealed elements retain their
values after the call,'' this would be stricter than necessary for many purposes.
For example, it arguably doesn't matter if {\tt f} overwrites {\tt main}'s data provided
that {\tt main} never reads that corrupted data.
This is the justification behind the lazy micro-policy~\cite{??}:
the monitor permits the caller to write into the
callee's frame, but taints that write so that the callee cannot read it later without
triggering a failstop. More generally, it is common to specify the behavior of a system,
including security properties, in terms of traces of observable events rather than details
of the machine state. For any such notion of observable behavior, we can consider
corruption by {\tt f} acceptable as long as it does not affect {\tt main}'s observable behavior.

To characterize this situation, we define a state element to be \emph{irrelevant}
if its value can be replaced with any other value without changing the observable behavior
of the machine. More precisely, for a set of elements \(\components\),
a pair of states \(\mach\) and \(\nach\)
are {\em \(\components\)-variants} if they agree on the value of every
element except those in \(\components\). The elements of \(\components\) are \emph{irrelevant}
in \(\mach\) if every variant state produces an equivalent trace
(see Section \ref{sec:events}); all other elements are \emph{relevant}.
We characterize \emph{observational integrity} as the property that
every relevant element that is \(\sealed\) under the callee's view is restored
to its original value at the return point.
